Supplmental Information : Understanding the nanoscale structure of hexagonal phase lyotropic liquid crystal membranes

### 2018-02-07 Ben_Manuscripts/structure_paper/Draft.tex
Updates to intro and supplemental

@@

%MRS5: a more compelling thesis is that we need membranes that are more selective than RO and NF.  
  RO and NF are finding applications beyond primarily water-based
  purifications. Recovery of valued products from complex aqueous and organic
  waste is of increasing interest. For example, flowback water (FW) produced
  during hydraulic fracturing is a complex wastewater full of potentially
  valuable dissolved organic compounds \cite{dischinger_application_2017}. There
  is increasing pressure to reuse hydraulic fracturing water rather than dispose
  of it in order to reduce social and envrionmental impacts as well as cost
  \cite{theodori_hydraulic_2014}. Rather than dispose of the wastestream
  generated in the recycling process, we can instead use highly selective
  membranes in order to successfully recover the useful compounds.

@@

%MRS5: maybe a more compelling thesis would be that LLC might have the capabilility to serve as selective membranes?
  Lyotropic liquid crystals (LLCs) are amphiphilic molecules that have the
  ability to self-assemble into porous nanostructures \cite{smith_ordered_1997}
  which can be crosslinked to create mechanically strong membrane films with
  pores on the order of 1 nm in diameter \cite{zhou_supported_2005}. Unlike
  commercial NF membranes, LLC membrane pores are uniform in size. Since LLC
  membranes lack a pore size distribution, they inherently exhibit high
  selectivity due to their strict molecular weight cut-off (MWCO). 
%MRS5: is there a citation for the strict MWCO? 
%MRS5: suggested change.
 Additionally, the LLCs chosen to make membranes for this application are salts, and therefore lead to 
Donnan exclusion of ions in solution.
 The membrane
  gains a net surface charge when counterions from head groups that line the pore
  walls escape into the feed solution in order to balance the gradients of
  concentration and electric potential. 
%This effect, which causes the membrane to
%  repel incoming charged species, is called Donnan exclusion

  \cite{donnan_theory_1995}.

  The feasiblity of LLC membranes has been demonstrated using LLCs that form
  the type 1 bicontinuous cubic (Q\textsubscript{I})
  \cite{hatakeyama_water_2011,hatakeyama_nanoporous_2010,carter_glycerol-based_2012},
  and the inverted hexagonal (H\textsubscript{II}) \cite{zhou_supported_2005}
  phases (See Figure~\ref{fig:bcc_v_hII}). When separating organic solutes from
  NaCl, Q\textsubscript{I} phase membrane filtration experiments have shown
%MRS5: for dashes between two numbers, use double -- for the correct length in the file. 
  selectivity 2--3 times higher than commercial RO and 6--12 times higher than
  commercial NF membranes \cite{dischinger_application_2017}.  When separating a
  series of various sized dyes, the H\textsubscript{II} phase membrane showed
  complete rejection of dyes bigger than 1.2 nm in size
  \cite{zhou_supported_2005}.

%MRS5: be more explicit about current efforts. Also use the
%opportunity of a thesis statement to convince the readership of
%something.  what do you want to convince them?
  We will focus our current efforts around the H\textsubscript{II} phase since
  its geometry should give optimal transport performance compared to the
  Q\textsubscript{I} phase. 
%MRS5: citation on this above?
  The H\textsubscript{II} phase forms at room
  temperature in the presence of c.a.~10 wt\% water and consists of hexagonally
  packed, straight, hydrophilic pore columns\cite{smith_ordered_1997}. In the
  absence of water, neat monomer will form the same hexagonal columnar structure
  which, in literature, has been referred to as the Col\textsubscript{h}
  thermotropic phase\cite{feng_scalable_2014} (See Figure~\ref{fig:hII_phase}).
  These hexagonal columnar phases represent the ideal geometry for membrane
  transport. 
%MRS5: maybe less about Q_I since it's not in this paper?  The introduction should be an introduction to the paper.
The most promising Q\textsubscript{I} phase for membrane
  applications forms at 70\degree C when monomer is mixed with c.a.~20 wt\%
  glycerol\cite{carter_glycerol-based_2012}. The Q\textsubscript{I} phase
  consists of a network of three dimensionally interconnected pores (See
  Figure~\ref{fig:bcc_phases}). This complex geometry effectively adds a degree
  of tortuosity to the membrane which prevents optimal through-plane transport.
  Despite the geometric differences between the Q\textsubscript{I} and
  H\textsubscript{II} phase, the chemical environment within the pores is
  expected to be similar, and either membrane should be highly selective.

>>>>

  Problem to be solved

@@

  We need membranes that are more selective than RO and NF. Recovery of valued
  products from complex aqueous and organic waste is of increasing interest. For
  example, flowback water (FW) produced during hydraulic fracturing is a complex
  wastewater full of potentially valuable dissolved organic compounds
  \cite{dischinger_application_2017}. There is increasing pressure to reuse
  hydraulic fracturing water rather than dispose of it in order to reduce social
  and envrionmental impacts as well as cost \cite{theodori_hydraulic_2014}.
  Rather than dispose of the wastestream generated in the recycling process, we
  can instead use highly selective membranes in order to successfully recover the
  useful compounds.

@@

  Crosslinked lyotropic liquid crystal (LLC) membranes may be capable of
  performing highly selective separations. LLCs are amphiphilic molecules that
  have the ability to self-assemble into porous nanostructures
  \cite{smith_ordered_1997} which can be crosslinked to create mechanically
  strong membrane films with pores on the order of 1 nm in diameter
  \cite{zhou_supported_2005}. Unlike commercial NF membranes, LLC membrane pores
  are uniform in size. Since LLC membranes lack a pore size distribution, they
  inherently exhibit high selectivity due to their strict molecular weight
  cut-off (MWCO) ~\cite{zhou_supported_2005}. Additionally, the LLCs chosen to
  make membranes for this application are salts, and therefore lead to Donnan
  exclusion of ions in solution. The membrane gains a net surface charge when
  counterions from head groups that line the pore walls escape into the feed
  solution in order to balance the gradients of concentration and electric
  potential \cite{donnan_theory_1995}.

  The feasiblity of LLC membranes has been demonstrated using LLCs that form
  the type 1 bicontinuous cubic (Q\textsubscript{I})
  \cite{hatakeyama_water_2011,hatakeyama_nanoporous_2010,carter_glycerol-based_2012},
  and the inverted hexagonal (H\textsubscript{II}) \cite{zhou_supported_2005}
  phases (See Figure~\ref{fig:bcc_v_hII}). When separating organic solutes from
  NaCl, Q\textsubscript{I} phase membrane filtration experiments have shown
  selectivity 2--3 times higher than commercial RO and 6--12 times higher than
  commercial NF membranes \cite{dischinger_application_2017}.  When separating a
  series of various sized dyes, the H\textsubscript{II} phase membrane showed
  complete rejection of dyes bigger than 1.2 nm in size
  \cite{zhou_supported_2005}.

%MRS5: be more explicit about current efforts. Also use the
%opportunity of a thesis statement to convince the readership of
%something.  what do you want to convince them?
  We will focus our current efforts around the H\textsubscript{II} phase since
  its geometry should give optimal transport performance compared to the
  Q\textsubscript{I} phase \cite{matyka_tortuosity-porosity_2008}. The
  H\textsubscript{II} phase forms at room temperature in the presence of c.a.~10
  wt\% water and consists of hexagonally packed, straight, hydrophilic pore
  columns\cite{smith_ordered_1997}. In the absence of water, neat monomer will
  form the same hexagonal columnar structure which, in literature, has been
  referred to as the Col\textsubscript{h} thermotropic
  phase\cite{feng_scalable_2014} (See Figure~\ref{fig:hII_phase}).  These
  hexagonal columnar phases represent the ideal geometry for membrane transport.

%MRS5: maybe less about Q_I since it's not in this paper?  The introduction should be an introduction to the paper.
%  The most promising Q\textsubscript{I} phase for membrane
%  applications forms at 70\degree C when monomer is mixed with c.a.~20 wt\%
%  glycerol\cite{carter_glycerol-based_2012}. The Q\textsubscript{I} phase
%  consists of a network of three dimensionally interconnected pores (See
%  Figure~\ref{fig:bcc_phases}). This complex geometry effectively adds a degree
%  of tortuosity to the membrane which prevents optimal through-plane transport.
%  Despite the geometric differences between the Q\textsubscript{I} and
%  H\textsubscript{II} phase, the chemical environment within the pores is
%  expected to be similar, and either membrane should be highly selective.

### 2018-02-07 Ben_Manuscripts/structure_paper/supplemental.tex
Updates to intro and supplemental

  \vspace{1em}
  \begingroup
	\fontsize{14pt}{12pt}\selectfont
	\textbf{Experimental correlation length of R-$\pi$}
  \endgroup

  \vspace{1em}
  Experimentalists did not measure the correlation length, so we estimated it
  from the raw data. We took a slice of the 2D WAXS data along the dashed line in
  Figure ~\ref{fig:waxs_dashed}. We removed background noise by subtracting the
  intensity far from the pattern (high $q$) uniformly. We used a least squares
  algorithm to fit a lorentzian curve to the data. Only values recorded above
  $q_z$ = 1.6 were considered for the fit in order to mitigate interference from
  R-alkanes (Figure~\ref{fig:correlation_length_exp}).  The full width at half
  maximum (FWHM) is related to the correlation length, L, by the relationship: $
  L = \frac{\pi}{FWHM} $. The error in the value is calculated as the square root
  of the diagonal entry of the covariance matrix calculated 
  % need citations and more on lorentzian parameters
  \begin{figure}[h]
  \centering
  \begin{subfigure}{0.45\textwidth}	
  	\includegraphics[width=\textwidth]{waxs_dashed.png}
	\caption{}~\label{fig:waxs_dashed}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
	\includegraphics[width=\textwidth]{Correlation_length_exp.png}
	\caption{}~\label{fig:correlation_length_exp}
  \end{subfigure}
  \caption{}~\label{fig:correlation}
  \end{figure}

  %Our model is insensitive to initial configuration if monomer placement is chosen within reason. 
  \begingroup
	\fontsize{14pt}{12pt}\selectfont
	\textbf{Initial Configuration Dependence}
  \endgroup

@@

		\item 55 \AA~: We tested a pore spacing of 55 \AA~ which is at
		a distance where monomers in each pore no longer intersect adjacent pores. Once
		the restraining potential reaches 56 KJ mol$^{-1}$ nm$^{-2}$, the pore spacing
		changes erratically until it begins to settle when the force constants are
		below 3 KJ mol$^{-1}$ nm$^{-2}$ (Figure~\ref{fig:p2p_55}). We reccomend
		avoiding a system such as this where vacuum gaps between pore columns are
		introduced unnecessarily. The system equilibrates to a pore spacing of 4.27
		$\pm$ 0.18 nm. The 1-2 pore spacing is the outlier in this case with a pore
                spacing of 3.99 \AA~ which distorts the hexagonal array.

@@

	  We tested 3 different pore radii, defined as shown in
	  Figure~\ref{fig:pore_radius_illustration}. For each system, we held the initial
	  pore spacing constant at 45 \AA~ and the distance between layers constant at
	  3.7 \AA. Equilibrated values of pore radii are presented in
	  Table~\ref{table:radii}. The pore radius in each simulation frame is calculated as the
	  average distance of all carbonyl carbons (See
	  Figure~\ref{fig:pore_radius_illustration}) from their associated pore center.
	  Statistics for the pore radii reported were generated from the timeseries
	  representing the average pore radius at each frame.  We detected equilibration 
	  using \texttt{pymbar.timeseries.detectEquilibration}. We calculated the average
	  and standard deviation of pore radii using only data points collected after
	  equilibration was detected.

@@

		\item 5 \AA~: We tested a pore radius of 5 \AA~because it is
		slightly larger than the equilibrated pore radius of the system simulated with
		an intial pore radius of 2.5 \AA. After being run through the full dry
		equilibration procedure, the average pore radius is 0.42 $\pm$ 0.01 \AA, which
		agrees with the 2.5 \AA~ configuration within error.

		\item 8 \AA~: The largest pore radius that can achieved before
		energy minimization becomes problematic is 8 \AA. One should use caution with
		such a structure because of the relatively large vacuum space that is created
		in the pore region of the initial configuration. After being run through the
		full dry equilibration procedure, we see a combination of cylindrical and
		slit-like pores (Figure~\ref{fig:slits}). We are wary of such a non-symmetric
		structure and choose not to use a pore radius of 8 \AA for our starting configurations.

@@

	  We tested 3 different initial layer spacings, defined as the distance
	  between the planes of aromatic rings in each layer. Systems built with layers
	  stacked 3.7 \AA~and 5 \AA~apart are discussed extensively in the main text. We
	  will focus the discussion here on systems built with layers stacked 10~\AA
	  apart.

@@

  Crosslinking of the the monomer Na-GA3C11 occurs by a UV initiated free
  radical polymerization (Figure~\ref{fig:xlink_mech}).  Head-to-tail addition
  takes place between terminal vinyl groups on each of the monomer tails.  We
  only considered head-to-tail addition since it is the dominant propagation mode
  in the real system.

>>>>

  \vspace{1em}
  \begingroup
	\fontsize{14pt}{12pt}\selectfont
	\textbf{\noindent Experimental correlation length of R-$\pi$}
  \endgroup

  \vspace{1em}
  Experimentalists did not measure the correlation length, so we estimated it
  from the raw data. We took a slice of the 2D WAXS data along the dashed line in
  Figure ~\ref{fig:waxs_dashed}. We removed background noise by subtracting the
  intensity far from the pattern (high $q$) uniformly. We used a least squares
  algorithm to fit a lorentzian curve to the data. Only values recorded above
  $q_z$ = 1.6 were considered for the fit in order to mitigate interference from
  R-alkanes (Figure~\ref{fig:correlation_length_exp}). The full width at half
  maximum (FWHM) is related to the correlation length, L, by the relationship: $
  L = \frac{1}{FWHM} $. The error in the value is calculated as the square root
  of the diagonal entry of the covariance matrix calculated 
  % need citations and more on lorentzian parameters
  \begin{figure}[h]
  \centering
  \begin{subfigure}{0.45\textwidth}	
  	\includegraphics[width=\textwidth]{waxs_dashed.png}
	\caption{}~\label{fig:waxs_dashed}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
	\includegraphics[width=\textwidth]{Correlation_length_exp.png}
	\caption{}~\label{fig:correlation_length_exp}
  \end{subfigure}
  \caption{}~\label{fig:correlation}
  \end{figure}
  \vspace{1em} \\
  %Our model is insensitive to initial configuration if monomer placement is chosen within reason. 
  \begingroup
	\fontsize{14pt}{12pt}\selectfont
	\textbf{Initial Configuration Dependence}
  \endgroup

@@

		\item 55 \AA~: We tested a pore spacing of 55 \AA~which is at
		a distance where monomers in each pore no longer intersect adjacent pores. Once
		the restraining potential reaches 56 KJ mol$^{-1}$ nm$^{-2}$, the pore spacing
		changes erratically until it begins to settle when the force constants are
		below 3 KJ mol$^{-1}$ nm$^{-2}$ (Figure~\ref{fig:p2p_55}). We reccomend
		avoiding a system such as this where vacuum gaps between pore columns are
		introduced unnecessarily. The system equilibrates to a pore spacing of 4.27
		$\pm$ 0.18 nm. The 1-2 pore spacing is the outlier in this case with a pore
                spacing of 3.99 \AA~ which distorts the hexagonal array.

@@

	  We tested 3 different pore radii, defined as shown in
	  Figure~\ref{fig:pore_radius_illustration}. For each system, we held the initial
	  pore spacing constant at 45 \AA~and the initial distance between layers
	  constant at 3.7 \AA. Equilibrated values of pore radii are presented in
	  Table~\ref{table:radii}. The pore radius in each simulation frame is calculated
	  as the average distance of all carbonyl carbons (See
	  Figure~\ref{fig:pore_radius_illustration}) from their associated pore center.
	  Statistics for the pore radii reported were generated from the timeseries
	  representing the average pore radius at each frame. We detected equilibration
	  using \texttt{pymbar.timeseries.detectEquilibration}. We calculated the average
	  and standard deviation of pore radii using only data points collected after
	  equilibration was detected.

@@

		\item 5 \AA~: We tested a pore radius of 5 \AA~because it is
		slightly larger than the equilibrated pore radius of the system simulated with
		an intial pore radius of 2.5 \AA. After being run through the full dry
		equilibration procedure, the average pore radius is 0.42 $\pm$ 0.01 \AA, which
		agrees with the 2.5 \AA~configuration within error.

		\item 8 \AA~: The largest pore radius that can achieved before
		energy minimization becomes problematic is 8 \AA. One should use caution with
		such a structure because of the relatively large vacuum space that is created
		in the pore region of the initial configuration. After being run through the
		full dry equilibration procedure, we see a combination of cylindrical and
		slit-like pores (Figure~\ref{fig:slits}). Measuring the pore radius of this
		system does not have a concrete meaning since slits do not have a single
		radius, but its calculated value is reported in Table ~\ref{table:radii}. We
		are wary of such a non-symmetric structure and choose not to use a pore radius
		of 8 \AA~for our starting configurations.

@@

	  We tested 3 different initial layer spacings, defined as the distance
	  between the planes of aromatic rings in each layer. Systems built with layers
	  stacked 3.7 \AA~and 5 \AA~apart are discussed extensively in the main text. We
	  will focus the discussion here on systems built with layers stacked 10 \AA~
	  apart.

@@

  \vspace{1em}
  Crosslinking of the the monomer Na-GA3C11 occurs by a UV initiated free
  radical polymerization (Figure~\ref{fig:xlink_mech}).  Head-to-tail addition
  takes place between terminal vinyl groups on each of the monomer tails.  We
  only considered head-to-tail addition since it is the dominant propagation mode
  in the real system.
